On the spatial aggregation of ecosystem condition indicators

### 2025-02-21 manuscript/manuscript.qmd
rm comments

SEEA EA includes principles for compiling accounts for ecosystem extent, ecosystem condition, and ecosystem services [@unitednations2024; -@unitednations2022].
We will focus on the SEEA EA condition accounts in this paper, although spatial aggregation is a relevant topic also for the other parts of the account, and to other types of indicator-based ecosystem assessments.
When EU and EEA countries are in 2026 required to report ecosystem condition accounts to Eurostat and the European Commision (?), these accounts will be spatially aggregated to national levels.
The point of having a robust methodological framework for the spatial aggregation of data and information is therefore quite pertinent.
<!--# thus invoking the issue of the ecological fallacy (Box 1), and many other potential pitfalls in the spatial aggregation of information. -->

@@

<!--# Let's illustrate this with an example using a hypothetical variable depicting wild animal densities. If the upper reference level (X100) for this variable is given by the assumed (or measured) animal densities under the reference condition, then that reference level needs to be assigned a relevant spatial scale, for example a region where the densities are calculated over. One will probably acknowledge the fact that animals move and distribute themselves non-randomly within these regions, and it therefore makes little sense to normalise the variable at a finer scale than region. Doing so would mean that regions perfectly aligned with the reference condition (in terms of animal densities with the region) will show as having sub-regions with sub-optimal condition, simply due to natural variation in where the animals are at any give time. Averaging (arguably the most common spatial aggregation function) the indicator values across sub-regions would not produce a regional indicator value that shows the region as being perfectly aligned with the reference condition, since truncation of variable values above X100 (assuming X100 > X0) means we are only aggregating the negative deviations from the reference level, and that very high animal densities in some sub-regions are not allowed to compensate for low values elsewhere. -->

@@

<!--# Subsequently, spatial aggregation is required when going from ecosystem assets to large spatial scale, ultimately the scale of the entire ecosystem accounting area (EAA).  -->

@@

For sparse data we are considering data where variable values are assigned to singular points or small areas.
In these cases, there will be a lot of area inside each ecosystem asset for which we have no direct information.
To assign variable value to all areas (as is required in the SEEA EA framework), we need to do some generalisations.
One can assign values to areas via an aggregation function used on data values that fall within a given ecosystem asset.
Many aggregation functions are available, such as sum, worst-rule, different percentiles, arithmetic, geometric or harmonic means[^2] [@unitednations2024 §5.54] .
When doing this aggregation, we need to account for the spatial structure (spatial correlations) of the data.
One way is to rely on design-based approaches (e.g. relying on random or balanced samples across the entire EEA), or using a step-wise aggregation via ecosystem assets or some smaller basal spatial units.
In our experience, the choice of which area units to use when doing this step-wise aggregation is often not reflected much upon.
In some cases the areal units are even decided politically, for example to administrative units, with little ecological justification for this choice.
However, this decision clearly affects the results of the spatial aggregation via a well-described (but often forgotten?) statistical problem referred to at the modifiable area unit problem (MAUP; Box 1).
It is important for decision makers to see that the information they are basing their decision on is robust and insensitive to aggregation methods [@vanbeurden1999].
The MAUP can, however, lead to inconsistencies between repeated or nested ECAs, leading to confusion and mistrust (@sec-comp).
One option to reduce this discrepancy and arbitrariness is to use ecologically motivated areal units in the step-wise aggregation such as homogeneous ecosystem areas [@vallecillo2022] or homogeneous impact areas [@kolstad_making_2024].

@@

```{=html}
<!-- \*Conclusion\*: Indicators should make intuitive sense to the readers.
However, there are sometimes good reasons for making things complicated. -->
```
:::

>>>>

SEEA EA includes principles for compiling accounts for ecosystem extent, ecosystem condition, and ecosystem services [@unitednations2024; -@unitednations2022].
We will focus on the SEEA EA condition accounts in this paper, although spatial aggregation is a relevant topic also for the other parts of the account, and to other types of indicator-based ecosystem assessments.
When EU and EEA countries are in 2026 required to report ecosystem condition accounts to Eurostat and the European Commision (?), these accounts will be spatially aggregated to national levels.
The point of having a robust methodological framework for the spatial aggregation of data and information is therefore quite pertinent.

@@

@@

@@

For sparse data we are considering data where variable values are assigned to singular points or small areas.
In these cases, there will be a lot of area inside each ecosystem asset for which we have no direct information.
To assign variable value to all areas (as is required in the SEEA EA framework), we need to do some generalisations.
One can assign values to areas via an aggregation function used on data values that fall within a given ecosystem asset.
Many aggregation functions are available, such as sum, worst-rule, different percentiles, arithmetic, geometric or harmonic means[^2] [@unitednations2024 §5.54] .
When doing this aggregation, we need to account for the spatial structure (spatial correlations) of the data.
One way is to rely on design-based approaches (e.g. relying on random or balanced samples across the entire EEA), or using a step-wise aggregation via ecosystem assets or some smaller basal spatial units.
In our experience, the choice of which area units to use when doing this step-wise aggregation is often not reflected much upon.
In some cases the areal units are even decided politically, for example to administrative units, with little ecological justification for this choice.
However, this decision clearly affects the results of the spatial aggregation via a well-described (but often forgotten?) statistical problem referred to at the modifiable area unit problem (MAUP; **Box 1**).
It is important for decision makers to see that the information they are basing their decision on is robust and insensitive to aggregation methods [@vanbeurden1999].
The MAUP can, however, lead to inconsistencies between repeated or nested ECAs, leading to confusion and mistrust (@sec-comp).
One option to reduce this discrepancy and arbitrariness is to use ecologically motivated areal units in the step-wise aggregation such as homogeneous ecosystem areas [@vallecillo2022] or homogeneous impact areas [@kolstad_making_2024].

@@

:::
